\chapter{Introduction}

\section{Preliminaries}

This dissertation investigates the cases of post-colonial, social and cultural implications that Arundhati Roy employs in the writing of her magnum opus, \emph{The God of Small Things}. This novel holds the title of being a critique of the post-colonial attitudes that exist in the middle-class milieu of the post-independent Indian society. This book that states the story of a fractured family from Ayemenem took the world by storm and went on to win the Booker Prize in the year 1997. The author created a brilliant post-colonial masterpiece which provides scholars and researchers with subjects that can be investigated and re-investigated under the post-colonial lens. The present dissertation is an attempt to club together issues such as subalternity, patriarchy, rootlessness, childhood trauma, oppression, cultural amnesia, multiculturalism, hybridity, post-colonial remembrance and other such myriad topics to prove the novel to be a classic post-colonial piece of work.

\section{Hypothesis}

It is being hypothesised that the characters in this novel continuously explore the consequences of being placed in a post-colonial world where they struggle to embrace the idea of cultural hybridity.

Infinite instances in the novel show glimpses of the post-colonial India, where ‘colonial’ does not only mean the natives being superseded by the white race, but also of assuming a power position by the natives themselves among their own people. 

\section{Aim}

This thesis aims at highlighting the post-colonial and cultural implications that exist in \emph{The God of Small Things}. The author, while penning down the story, uses many distinguishing features of post-colonial theory such as the issues of mimicry, hybridity, identity, untouchability and oppressive superiority to draw a portrait of the post-colonial, chaotic India.

\section{Objectives}

\begin{enumerate}
  \item To expose the ideology of patriarchal oppression as well as the hegemony of socio-political systems.
  \item To interpret the myriad ways in which the dominated classes try to safeguard themselves against their potential oppressors on the subversion of their human rights.
  \item To witness the debunking of moral and political ethics to give way to oppression and violence.
  \item To show the after effects of appropriation of the western values by the mimicking natives.
  \item To represent the state of the marginalized or the less empowered and the exposition of the hypocritical power structure.
  \item To analyse the acceptance of hybridity as an effective solution to post-colonial angst.
\end{enumerate}


\section{Methodology and Technique of Study}

\emph{The God of Small Things} has been selected as the primary source of study. The text has been analysed from the post-colonial perspective. This analysis is being made after the primary reading of the novel that depicts innumerable instances of the theory being juxtaposed in its layered storytelling. Moreover, notable work and theories of renowned post-colonial authors have been taken into consideration to justify the analysis.

Throughout the past few decades, post-colonial texts have risen to prominence in India following its independence. Such texts not only study the relationship between the now-gone coloniser and colonised but also the effects that it has on the people of the newly independent country. \emph{The God of Small Things} is a striking example of such a discourse that explores the post-colonial India, during the 1960s through the lives of the fraternal twins, Rahel and Estha, and their family in Ayemenem in south India.

The postcolonial study, in this novel, especially deals with the effects of colonization on the characters in its aftermath. Most of the characters are hybrids, and most of the cultural images used are foreign, yet the characters who are native Indians, consider such images to be an intrinsic part of their lifestyle. Therefore, the present study, as mentioned in the aim, makes an effort to highlight the post-colonial perspectives in the novel.

\section{The Significance of the Study}

This research not only adds to the post-colonial analysis of the natives in newly independent India but also depicts the subtle cast biases and issues such as untouchability that exist in the new post-independent India. Also, scholars have hardly studied the characters as hybrid mimics of the now gone westerners. It is fascinating to explore such aspects of the novel too wherein the natives are read as people adopting or mimicking the positions of superiority to dominate on their less powerful brethren. The study finally reveals the truth of all of us being a conscious or an unconscious hybrid of our past. Accepting this truth can be a relevant solution for achieving peace with oneself and one’s surrounding.

\section{Postcolonial Prospects in The God of Small Things}

The God of Small Things undoubtedly presents and reflects the issues of the post-colonial period. Arundhati Roy was born, grew up and was educated in India. Roy in her celebrated novel The God of Small Things tells the story of a Syrian Christian family living in the southern province of Kerala, India. The novel critiques the aftermath of the colonial phase of India, the working of colonial forces even after the independence of the country, nevertheless this allows the possibility of correspondence with the West.  Roy employs a similar theoretical perspective that reverses the identity of the ‘Other’. She depicts the post-colonial ‘Other’ in newer forms of cultural, patriarchal, and political oppression that result from colonization.

The plot revolves around a family that includes few members such as Pappachi Kochamma, father of Ammu, who had worked under the imperial power as an entomologist and had retired by the time the story of the novel begins. His retirement also marked the end of the imperial power after which he returns to his hometown, Ayemenem, in Kerala along with his family. In this novel, this incident is recounted in hindsight when Pappachi’s daughter Ammu too returned to live with her parents in Ayemenem, along with her two children, Estha and Rahel, after walking out of an unhappy marriage by divorcing her husband. Pappachi’s son, Chacko had gone to England, to study at Oxford University wherein he meets his prospective wife, Margaret. They get married in England and have a daughter named Sophie Mol, but unfortunately, their marriage comes to an end within a year. This tragedy brings Chacko back to his parents’ house in Ayemenem, leaving his daughter and divorced wife behind.

The main plot that connects all the other plots floating in the novel begins when Sophie Mol, along with her mother, comes to visit her father in India, and the consequent catastrophic event of her drowning which augments the dismantling of an already disintegrated family. The incident of Mol’s drowning entails exceedingly to the immense amount of grief and bewilderment that the family experienced. Their misery is further elevated when they become aware of Ammu and Velutha’s love-affair. Velutha is the family’s carpenter and an untouchable of a lower caste. The deeply entrenched notion of segregation and discrimination that the caste system propagated in India, prohibited close contact with an untouchable. Therefore, quite obviously, having an affair with an untouchable was altogether a forbidden terrain.

In her attempts at interaction and intimate contact with Velutha, Ammu transgresses the rigid boundaries of caste. This causes the family to fall apart and consequently leads to the separation of Ammu’s twins, Estha and Rahel from each other. The seemingly simplistic plot, when unfolds in a gradual progression via its peculiar narration, lays bare the deeply rooted post-colonial discourse and various post-colonialist features that are intricately working within the text. \parencite{Nayar}

\section{A Perspective on the Author}

Suzanna Arundhati Roy’s individual experiences are as alluring as her narrative. Roy came of age in Kottayam, a province in Kerala, which also served as the setting for her novel. She is the daughter of a Hindu Bengali father and a Christian mother. She led an unconventional life since early adolescence. Roy grew up with her mother in Ayemenem after her parents got divorced in her childhood. One can identify her autobiographical adoration for Ayemenem all through the novel. Once school got over, she enrolled herself in Delhi’s architectural college, similar to what Rahel does.

She confessed in an interview that this experience did shape up a large portion of her aspiration to write a novel. She began to write scripts for television and won a national award too for some of her work. She chose to write privately after an escalating controversy regarding the movie Bandit Queen, the screenplay of which was drafted by her. Later she went ahead to work as an aerobics instructor to take a breather. She mentions in her interviews that it was her inner calling that inspired her to write The God of Small Things that chartered her career as an established writer.

Her style of writing displays the literary texture of Ayemenem. On being asked why she chose Kerala as her setting, she says, ``it was the only place in the world where religions coincide, there's Christianity, Hinduism, Marxism and Islam and they all live together and rub each other down\ldots I was aware of the different cultures when I was growing up, and I'm still aware of them now. When you see all the competing beliefs against the same background, you realize how they all wear each other down.'' \parencite[p.~36]{TalwarShashi}  Roy later became a political activist and apart from writing fiction, continues to work at the literary front too. She appeared as a distinguished author after her monumental work The God of Small Things hit the market. It is an unprecedented novel in manner and matter. It was her linguistic ingeniousness that brought her the prestigious Booker Prize. Her biography is as enthralling as her novel.

\section{What Does the Text Carry?}

A close reading of the novel reveals that the entire development of the book has been unfolded in gist in the first chapter itself. She declares, ``I would start some where and I’d colour in a bit, and then I would deeply stretch back and then stretch forward. It was like designing an intricate balanced structure.'' \parencite{TalwarShashi}. The novel consists twenty-one chapters sewn together in the form of fragmented memories and ruptured imaginings, being narrated by the thirty-one-year-old Rahel who returns to her hometown after ages. The text magically interweaves the past and the present through the use of flashback consciousness. The Booker commendation focuses on Roy’s use of history which is undoubtedly a paramount factor in post-colonial writings. It says, ``With extraordinary linguistic inventiveness Roy funnels the history of South India through the eyes of seven-year-old twins.'' (The Week, 47).


About narrating the novel or the world of adults from a child’s perspective, she says it was a conscious decision as she had an unprotected childhood. She perhaps wanted to intensify the drama by depicting it through a young narrator’s vision. She explains in one of her interviews:

\begin{quote}
  “I had an unprotected childhood\ldots Two things happen. You grow up quickly. And when you become an adult there is a part of you that remains a child. So the communication between you and your childhood remains open.” \parencite{Abraham}
\end{quote}

Penning down a novel from a child’s perspective should only associate it with more authenticity at a symbolic level because children are believed to be authentic regarding conveying their emotions. It attempts to break many taboos still persistent in a post-colonial, post independent world.

\section{Major Postcolonial Themes in the Novel}

The novel aims to raise social consciousness by exposing the injustice and tyranny inflicted upon the untouchables. It focuses on the tribulations and insults experienced by the defenceless and the deserted in the police custody. It features the class distinction prevailing in the society. Roy portrays the harsh irony of the male’s domination over the female. She shows how the male not only objectifies or belittles the female but, on the contrary, can also worship her to possess. What’s problematic is that the men in the novel can rarely consider women to be their equals. The novel is a satire on religion, power and politics too. There are allusions to dormant characters, like Velutha’s grandfather, who have undergone a change of religion just for avoiding the evils of untouchability. Similarly, the police act as a plaything in the hands of politicians. Most of the times the politicians act against the motto that they stand for.

The novelist has eruditely put in her work two significant metaphors, ‘Laltain’ and ‘Mombatti’. While both can be classified together as being used for similar purpose, one has a well-covered protective layer while the other burns defenceless. A sudden surge of wind is enough to succumb the latter. Through such symbolic connotations, Roy kindles the reader’s sympathy for the defenceless. Roy as a Nineties’ writer is an archetype of the present scenario of Indian writing in English. She has come through as a relentless disapprover of the process of clutching on to a traditional Indian way of living life. She is postmodern in her thoughts and applies that to her art.

This novel of struggle and protest against the downtrodden includes Dalits and women in its sphere. Their lives are intertwined with patriarchal norms and caste corruption. It highlights the exploitative nature of the norms that smother the Dalit’s aspirations. Roy constructed the structure of the novel keeping the suffering of the women as a pivotal point. The plot has been set in 1960’s post-independent India to highlight the ironical position into which the subjugated classes are put after freedom. Outwardly the country seems to be governed by modern democratic principles, but internally the rule is imperialistic. Externally the façade of communistic harmony reigns in the public sphere, but man is a capitalist, imperialist and a patriarch, intrinsically. The consequences of such a system are obvious, the repression of the already downtrodden. The novel is a representation of a continuous struggle between the worst transgressors and the authoritarians that keep the dynamism alive.