\newpage
\addcontentsline{toc}{chapter}{ACKNOWLEDGEMENTS}
\begin{center}
    \LARGE ACKNOWLEDGEMENTS
\end{center}
This  work  was  carried  out  during  the  years  2016-2018  at  the  Institute  of  Advanced  Studies  in  English  (IASE),  Pune.  

I  owe  my  deepest  gratitude  to  my  supervisor,  Director  and  Dean  of  IASE,  Professor  (Dr)  Ashok  Thorat.  Without  his  continuous  optimism,  encouragement,  enthusiasm  and  support,  this  dissertation  would  hardly  have  seen  the  light  of  the  day.  During  the  most  difficult  times,  he  provided  me  with  moral  support  and  the  independence  to  experiment.  Sir,  I  owe  it  all  to  you.

I  am  indebted  to  all  my  professors  at  IASE  who  build  all  our  batch  mates  into  sincere  and  accomplished  research  scholars.  This  dissertation  is  not  only  an  effort  at  the  keyboard  but  gratitude  to  their  candid  coaching.

I  must  also  acknowledge  the  rigorous  support  provided  by  the  ever  pleasant,  Ms  Triveni  Shinde,  Administrator  (IASE),  during  the  formative  years  of  every  student  at  the  institute,  at  all  levels.  It  will  not  be  an  exaggeration  to  mention  that  she  has  pushed  many  students,  including  myself,  to  achieve  their  limits  through  her  administrative  and  emotional  support.  My  batch  mates'  invaluable  comments  during  the  brainstorming  group  sessions  served  as  warm  encouragement  in  the  designing  of  this  work.  

Finally,  I  must  convey  my  heartfelt  indebtedness  to  my  family  for  providing  me  with  constructive  criticism  and  extraordinary  support  throughout  my  academic  years,  and  through  the  process  of  researching  and  penning  down  this  dissertation.  Without  their  persistent  help,  this  dissertation  would  not  have  seen  the  light  of  the  day.  Thank  you.
